---
title: "SimonHeinberg_A02_CodingBasics.Rmd"
author: "Simon_Heinberg"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.
```{r}
sequence30 <- (seq(0, 30, 3)) #I created an object named 'sequence30' that displays a sequence of numbers from 0 to 30 by intervals of 3

```

2.  Compute the mean and median of this sequence.

```{r}
meanfunction <- mean(sequence30) 

medianfunction <- median(sequence30) #I created an object named 'meanfunction' that calculates the mean of 'sequence30' and an object named 'medianfunction' that calculates the median of 'sequence30

```


3.  Ask R to determine whether the mean is greater than the median.
```{r}
meanfunction > medianfunction #I asked R to give a 'TRUE' or 'FALSE' answer to whether the output of 'meanfunction' is greater than the 'medianfunction'
```

4.  Insert comments in your code to describe what you are doing.

I created an object named 'sequence30' that displays a sequence of numbers from 0 to 30 by intervals of 3



I created an object named 'meanfunction' that calculates the mean of 'sequence30' and an object named 'medianfunction' that calculates the median of 'sequence30'


I asked R to give a 'TRUE' or 'FALSE' answer to whether the output of 'meanfunction' is greater than the 'medianfunction'


## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.


```{r}
students <- c("jill", "jack", "james", "joy") 

scores <- c(35, 90, 72, 95) 

pass_fail <- scores>50 

```


6.  Label each vector with a comment on what type of vector it is.
```{r}
students #vector of student names

scores #vector of student scores

pass_fail #vector of whether or not they have passed the test
```

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.
```{r}

df <-data.frame(students, scores, pass_fail)


```

8.  Label the columns of your data frame with informative titles.

```{r basics 2}

names(df) <-c("studentnames", "studentscores", "did_the_student_pass") 

```

After failing to find a function for renaming the columns of a dataframe, I asked chatGPT which function I should use to rename a dataframe. I found the names(df) function through chatGPT.

9.  QUESTION: How is this data frame different from a matrix?

> Answer: The data frame is different from a matrix because it has three types of data, whereas a matrix only has one type of data.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 
```{r}

  pass_fail_function <-function(x) {ifelse(x>50, print("pass"), print ("fail"))}

```

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
pass_fail_output <- pass_fail_function(scores)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:'ifelse' worked since it ran the function for all four objects in the vector, calculating the expression x>50 for each object in the vector. 'if' and 'else' should also work since it is the longer notational form of the 'ifelse' function.
